Dysregulation of Cerebral Hemoglobin Content in Malawian Children with Malaria

### 2021-05-13 figures.Rmd
add variables to table 1

# Trim for variables of interest (VOI)
df_hbox_trim <- df_hbox %>% 
  dplyr::select(c(Subject.ID..NIAID., Status, Glucose, Hematocrit, Lactate, Temperature,
                  Arginine.umol.L, Haptoglobin..mg.dl., Hemoglobin..uM., Whole.Blood.Nitrite, 
                  cerebral.hbdeox, cerebral.hbdeox.alpha, cerebral.hbox, cerebral.hbox.alpha,
                  cerebral.sat, cerebral.sat.alpha, cerebral.thc, cerebral.thc.alpha, cerebral.thc.norm.std)) %>%

  dplyr::rename("subject_id" = "Subject.ID..NIAID.", "Arginine (umol/L)" = "Arginine.umol.L", "Haptoglobin (mg/dL)" = "Haptoglobin..mg.dl.", 
                "Hemoglobin (uM)" = "Hemoglobin..uM.", "WB_Nitrite" = "Whole.Blood.Nitrite", "DeoxyHb" = "cerebral.hbdeox", 
                "DeoxyHb_alpha" = "cerebral.hbdeox.alpha", "OxyHb" = "cerebral.hbox", "OxyHb_alpha" = "cerebral.hbox.alpha", 
                "O2sat" = "cerebral.sat", "O2sat_alpha" = "cerebral.sat.alpha", "THC" = "cerebral.thc", 
                "THC_alpha" = "cerebral.thc.alpha", "THC_sd" = "cerebral.thc.norm.std")

@@

# pivot long
df_hbox_long <- df_hbox_trim %>% pivot_longer(cols = 3:ncol(.), names_to = "measurement", values_to = "value")

@@

# Age, hematocrit, and lactate
df_hbox %>% 
  group_by(Status) %>% 
  summarise(`Age (years)` = median(Age, na.rm = TRUE), age_range = range(Age, na.rm = TRUE), 
            `Hematocrit (%)` = median(Hematocrit, na.rm = TRUE), hem_range = range(Hematocrit, na.rm = TRUE),
            `Lactate (mmol/L)` = median(Lactate, na.rm = TRUE), lactate_range = range(Lactate, na.rm = TRUE),
            `Arginine (umol/L)` = median(Arginine.umol.L, na.rm = TRUE), arg_range = range(Arginine.umol.L, na.rm = TRUE),
            `Temperature (°C)` = median(Temperature, na.rm = TRUE), temp_range = range(Temperature, na.rm = TRUE)) %>% 
  mutate_if(is.numeric, round, 1) %>%

  mutate(`Age (years)` = paste0(`Age (years)`, " (", age_range[1], ", ", age_range[2], ")"),
         `Hematocrit (%)` = paste0(`Hematocrit (%)`, " (", hem_range[1], ", ", hem_range[2], ")"),
         `Lactate (mmol/L)` = paste0(`Lactate (mmol/L)`, " (", lactate_range[1], ", ", lactate_range[2], ")"),
         `Arginine (umol/L)` = paste0(`Arginine (umol/L)`, " (", arg_range[1], ", ", arg_range[2], ")"),
         `Temperature (°C)` = paste0(`Temperature (°C)`, " (", temp_range[1], ", ", temp_range[2], ")")) %>% 
  dplyr::select(-c(age_range, hem_range, lactate_range, arg_range, temp_range)) %>% 
  unique() %>% 
  t() %>% 
  row_to_names(row_number = 1) %>% 
  as_tibble(rownames = "variable") %>% 
  add_row(df_status, .before = 1) %>% 
  add_row(df_sex, .before = 2) %>%

  dplyr::rename(" " = "variable",
                "Healthy volunteers (HV)" = "HV",
                "Uncomplicated malaria patients (UM)" = "UM",
                "Cerebral malaria patients (CM)" = "CM") %>% 
  knitr::kable(align = rep('c', 4))

@@

### Figure 2: Random forest modeling using all outcomes and predictors

#### A. PCA
```{r fig2a}

# convert tibble to df
df_hbox_impute_df <- df_hbox_impute %>% 
  dplyr::select(-c(subject_id, Status)) %>% 
  dplyr::rename("Whole blood nitrite" = "WB_Nitrite", "THC sd" = "THC_sd", "THCα" = "THC_alpha", 
                "Deoxy Hb" = "DeoxyHb", "Oxy Hb" = "OxyHb", "O2 sat" = "O2sat", "O2 sat α" = "O2sat_alpha",
                "Deoxy Hb α" = "DeoxyHb_alpha", "Oxy Hb α" = "OxyHb_alpha",
                "Hemoglobin" = "Hemoglobin (uM)", "Haptoglobin" = "Haptoglobin (mg/dL)", "Arginine" = "Arginine (umol/L)") %>% 
  as.data.frame(df_hbox_impute) 
rownames(df_hbox_impute_df) <- df_hbox_impute$subject_id

# compute PCA
my_pca1 <- prcomp(df_hbox_impute_df, scale = TRUE)

# define groups
groups1 <- as.factor(df_hbox_impute$Status)

# Eigenvalues
pca_eig_val1 <- get_eigenvalue(my_pca1)

# extract variables
df_vars1 <- my_pca1$rotation %>% as_tibble() %>% mutate(variable = rownames(my_pca1$rotation))

# extract patients
df_pnts1 <- my_pca1$x %>% as_tibble() %>% mutate(group = groups1)

  # plot patients
  ggplot() +
  geom_point(data = df_pnts1, 
             mapping = aes(x = PC1, y = PC2, color = group), 
             size = 3) +
  stat_conf_ellipse(data = df_pnts1, 
                    mapping = aes(x = PC1, y = PC2, color = group, fill = group),
                    level = 0.95, npoint = 100, bary = TRUE, alpha = 0.1, geom = "polygon") +
  stat_mean(data = df_pnts1, 
            mapping = aes(x = PC1, y = PC2, color = group, shape = group),
            na.rm = FALSE) +

  # plot variables
  geom_point(data = df_vars1, aes(x = PC1*10, y = PC2*10)) +
  geom_text_repel(data = df_vars1, aes(x = PC1*10, y = PC2*10, label = variable, size = 10), show.legend = FALSE) +
  geom_segment(data = df_vars1, aes(x = 0, y = 0, xend = PC1*10, yend = PC2*10), arrow = arrow()) +

  # design
  scale_color_manual(values = my_col) +
  scale_fill_manual(values = my_col) +
  xlab(paste0("PC1: ", round(pca_eig_val1$variance.percent[1], 2), "% of variance")) +
  ylab(paste0("PC2: ", round(pca_eig_val1$variance.percent[2], 2), "% of variance")) +

#### B. Confusion matrix
```{r fig2b_prep, eval = FALSE}

# Data
df_hbox_model1 <- df_hbox_trim %>% 
  dplyr::select(-subject_id) %>% 
  dplyr::rename("Whole blood nitrite" = "WB_Nitrite", "THC sd" = "THC_sd", "THCα" = "THC_alpha", 
                "Deoxy Hb" = "DeoxyHb", "Oxy Hb" = "OxyHb", "O2 sat" = "O2sat", "O2 sat α" = "O2sat_alpha",
                "Deoxy Hb α" = "DeoxyHb_alpha", "Oxy Hb α" = "OxyHb_alpha",
                "Hemoglobin" = "Hemoglobin (uM)", "Haptoglobin" = "Haptoglobin (mg/dL)", "Arginine" = "Arginine (umol/L)")

# Build a model

  # resamples (not a lot of data, need to use cross validation)
  set.seed(123)
  hbox_boot1 <- bootstraps(df_hbox_model1, strata = Status)

  # Create recipe
  ranger_recipe <- 
    recipe(formula = Status ~ ., data = df_hbox_model1) %>%

    # BoxCox transformation to normality
    step_BoxCox(all_predictors(), -all_outcomes()) %>%

    # remove variables with non-zero variance
    step_nzv(all_predictors(), -all_outcomes()) %>%

    # impute missing data using K-nearest neighbors
    step_knnimpute(all_predictors(), -all_outcomes())

  # Random forest model specifications (set for tuning)
  ranger_spec <- 
    rand_forest(mtry = tune(), min_n = tune(), trees = 1000) %>% 
    set_mode("classification") %>% 
    set_engine("ranger")

  # Create workflow
  ranger_workflow <- 
    workflow() %>% 
    add_recipe(ranger_recipe) %>% 
    add_model(ranger_spec)

# Tune model

  # Initial tune on bootstraps
  set.seed(84374)
  ranger_tune <-
    tune_grid(ranger_workflow, 
              resamples = hbox_boot1, 
              grid = 11)

# Finalize model

  final_rf <- ranger_workflow %>% 
    finalize_workflow(select_best(ranger_tune, metric = "accuracy"))

  final_rs_res <- final_rf %>% 
    fit_resamples(resamples = hbox_boot1,
                  metrics = metric_set(roc_auc, accuracy, sens, spec),
                  control = control_resamples(save_pred = TRUE))

# save objects so model doesn't rerun every time you knit
  save(df_hbox_model1, hbox_boot1, ranger_recipe, ranger_spec, ranger_tune, final_rs_res, file = "model1.Rdata")

```{r fig2b_plot}

final_res_conf_mat <- final_rs_res %>% conf_mat_resampled() 
final_res_conf_mat %>% 
  group_by(Truth) %>% 
  summarise(total_true = sum(Freq)) %>% 
  right_join(final_res_conf_mat) %>% 
  mutate(`Percent occurence` = round(Freq/total_true*100, 2)) %>%

  # plot
  ggplot(aes(x = Truth, y = Prediction, fill = `Percent occurence`, label = `Percent occurence`)) +
  geom_tile() +
  geom_text(size = 8) +
  scale_fill_gradient2(low = "#4575b4", high = "#d73027") +

#### C. Variable importance
```{r fig2c}

# new specification for ranger model
imp_spec <- ranger_spec %>% 
  finalize_model(select_best(ranger_tune, metric = "accuracy")) %>% 
  set_engine("ranger", importance = "permutation")

# plot importance 
workflow() %>% 
  add_recipe(ranger_recipe) %>% 
  add_model(imp_spec) %>% 
  fit(df_hbox_model1) %>% 
  pull_workflow_fit() %>% 
  vip(aesthetics = list(alpha = 0.8, fill = "midnightblue")) +

  theme_bw() +
  theme(axis.text = element_text(size = 20),
        axis.title = element_text(size = 25))

**Figure 2.** Random forest modeling demonstrates a high degree of discriminatory power between healthy patients and patients with malaria (uncomplicated and cerebral) using primarily on hematological parameters. **(A)** Principal component analysis (PCA) of study patients using hematological and NIRS data. Variables are plotted as vectors in the direction and magnitude of the PC to which they contribute. Color indicates patient status, ellipses show median and 95% confidence interval on the cluster. **(B)** Confusion matrix representing the results of random forest modeling averaged across 25 bootstrap resamples. Freq indicates the average number of times that the event (shown by the x- and y-axes) occurred. **(C)** Relative contribution of the variables to the discriminatory power of the model (showing top 10 only).

@@

### Figure 4: Relationship between THCα and THC sd

```{r fig4}

cm_model <- lm(THC_sd ~ THC_alpha, data = df_hbox_log_voi %>% filter(Status == "CM"))
um_model <- lm(THC_sd ~ THC_alpha, data = df_hbox_log_voi %>% filter(Status == "UM"))

df_hbox_log_voi %>% dplyr::rename("THC sd" = "THC_sd", "THCα" = "THC_alpha") %>%

  ggplot(aes(x = THCα, y = `THC sd`, color = Status, fill = Status)) +
  geom_point(size = 3, alpha = 0.7) +
  geom_smooth(method = "lm", se = FALSE) +

  geom_text(label = paste0("y = ", 
                           round(cm_model$coefficients[[1]], 2), 
                           " + ", 
                           round(cm_model$coefficients[[2]], 2), 
                           "x ; R^2 = ", 
                           round(summary(cm_model)$adj.r.squared, 2)),
            x = 1.3, y = 0.55,
            color = "#00BFC4",
            size = 6) +

    geom_text(label = paste0("y = ", 
                           round(um_model$coefficients[[1]], 2), 
                           " + ", 
                           round(um_model$coefficients[[2]], 2), 
                           "x ; R^2 = ", 
                           round(summary(um_model)$adj.r.squared, 2)),
            x = 1.3, y = 0.6,
            color = "#F8766D",
            size = 6) +

**Figure 4.** Scatter plot demonstrating the relationship between THCα and THC sd.

>>>>

# Trim for variables of interest (VOI)
df_hbox_trim <- df_hbox %>% 
  dplyr::select(c(Subject.ID..NIAID., Status, Glucose, Hematocrit, Lactate, Temperature,
                  Arginine.umol.L, Haptoglobin..mg.dl., Hemoglobin..uM., 
                  BP.Diastolic, BP.Systolic, O2.Sat, RR, HR,

                  cerebral.hbdeox, cerebral.hbdeox.alpha, cerebral.hbox, cerebral.hbox.alpha,
                  cerebral.sat, cerebral.sat.alpha, cerebral.thc, cerebral.thc.alpha)) %>%

  dplyr::rename("subject_id" = "Subject.ID..NIAID.", 
                "Arginine (umol/L)" = "Arginine.umol.L", "Haptoglobin (mg/dL)" = "Haptoglobin..mg.dl.", 
                "Hemoglobin (uM)" = "Hemoglobin..uM.", "Lactate (mmol/L)" = "Lactate", "Hematocrit (%)" = "Hematocrit",
                "Systolic BP" = "BP.Systolic", "Diastolic BP" = "BP.Diastolic", "O2 sat" = "O2.Sat", 
                "Respiration rate" = "RR", "Heart rate" = "HR",

                "Deoxy Hb" = "cerebral.hbdeox", "Deoxy Hb α" = "cerebral.hbdeox.alpha", 
                "Oxy Hb" = "cerebral.hbox", "Oxy Hb α" = "cerebral.hbox.alpha",  
                "Cerebral O2 sat" = "cerebral.sat", "Cerebral O2 sat α" = "cerebral.sat.alpha", 
                "THC" = "cerebral.thc", "THCα" = "cerebral.thc.alpha")

# Remove patients labeled as "UM" that instead meet the criteria for severe malaria

  # remove UM patient with lactate > 5 
  df_hbox_trim <- df_hbox_trim %>% dplyr::filter(!(Status == "UM" & `Lactate (mmol/L)` > 5))

  # All UM patients have hematocrit > 15%
  # df_hbox_trim %>% dplyr::filter(Status == "UM"  & `Hematocrit (%)` <= 15)

@@

# pivot long
df_hbox_long <- df_hbox_trim %>% pivot_longer(cols = 3:ncol(.), names_to = "variable", values_to = "value")

@@

# Age, hematocrit, and lactate
df_hbox %>%

  group_by(Status) %>% 
  summarise(`Age (years)` = median(Age, na.rm = TRUE), age_range = range(Age, na.rm = TRUE),

            `Temperature (°C)` = median(Temperature, na.rm = TRUE), temp_range = range(Temperature, na.rm = TRUE),
            `Heart rate` = median(HR, na.rm = TRUE), hr_range = range(HR, na.rm = TRUE),
            `Systolic BP` = median(BP.Systolic, na.rm = TRUE), sys_range = range(BP.Systolic, na.rm = TRUE),
            `Diastolic BP` = median(BP.Diastolic, na.rm = TRUE), dia_range = range(BP.Diastolic, na.rm = TRUE),
            `Respiration rate` = median(RR, na.rm = TRUE), rr_range = range(RR, na.rm = TRUE),
            `O~2~ saturation` = median(O2.Sat, na.rm = TRUE), o2_range = range(O2.Sat, na.rm = TRUE),

            `Hematocrit (%)` = median(Hematocrit, na.rm = TRUE), hem_range = range(Hematocrit, na.rm = TRUE),
            `Lactate (mmol/L)` = median(Lactate, na.rm = TRUE), lactate_range = range(Lactate, na.rm = TRUE),
            `Arginine (μmol/L)` = median(Arginine.umol.L, na.rm = TRUE), arg_range = range(Arginine.umol.L, na.rm = TRUE)) %>% 
  mutate_if(is.numeric, round, 1) %>%

  mutate(`Age (years)` = paste0(`Age (years)`, " (", age_range[1], ", ", age_range[2], ")"),

         `Temperature (°C)` = paste0(`Temperature (°C)`, " (", temp_range[1], ", ", temp_range[2], ")"),
         `Heart rate` = paste0(`Heart rate`, " (", hr_range[1], ", ", hr_range[2], ")"),
         `Systolic BP` = paste0(`Systolic BP`, " (", sys_range[1], ", ", sys_range[2], ")"),
         `Diastolic BP` = paste0(`Diastolic BP`, " (", dia_range[1], ", ", dia_range[2], ")"),
         `Respiration rate` = paste0(`Respiration rate`, " (", rr_range[1], ", ", rr_range[2], ")"),
         `O~2~ saturation` = paste0(`O~2~ saturation`, " (", o2_range[1], ", ", o2_range[2], ")"),

         `Hematocrit (%)` = paste0(`Hematocrit (%)`, " (", hem_range[1], ", ", hem_range[2], ")"),
         `Lactate (mmol/L)` = paste0(`Lactate (mmol/L)`, " (", lactate_range[1], ", ", lactate_range[2], ")"),
         `Arginine (μmol/L)` = paste0(`Arginine (μmol/L)`, " (", arg_range[1], ", ", arg_range[2], ")")) %>% 
  dplyr::select(-c(age_range, temp_range, hr_range, sys_range, dia_range, rr_range, o2_range, hem_range, lactate_range, arg_range)) %>%

  unique() %>% 
  t() %>% 
  row_to_names(row_number = 1) %>% 
  as_tibble(rownames = "variable") %>% 
  add_row(df_status, .before = 1) %>% 
  add_row(df_sex, .before = 2) %>%

  dplyr::rename(" " = "variable",
                "Healthy volunteers (HV)" = "HV",
                "Uncomplicated malaria (UM) patients" = "UM",
                "Cerebral malaria (CM) patients" = "CM") %>% 
  knitr::kable(align = rep('c', 4))

@@

@@
